---
title: "Border Crossing Entry Data"
---

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
library(ggthemes)
library(scales)
library(plotly)
```

```{r}
#| label: view

show <- read_csv("01-unclean/Border_Crossing_Entry_Data.csv")

show
```


```{r}
#| label: import

border <- read_csv("01-unclean/Border_Crossing_Entry_Data.csv") |>
  clean_names() |>
  mutate(
    border_year = my(date),
    year = year(border_year)
  ) |>
  select(state, border, year, measure, value)

border
```
### Cleaning

```{r}
#| label: cleaning

border |>
  write_rds("02-clean/border.rds")
```

## Analysis and Visuals

### Numbers by State

```{r}
#| label: clean import

clean_border <- read_rds("02-clean/border.rds")
```

```{r}
clean_border |>
  count(year)
```


```{r}
#| label: state values

border_state <- clean_border |>
  filter(
    year == 2024
  ) |>
  group_by(state, year) |>
  summarize(
    border_traffic = sum(value)
  ) |>
  arrange(desc(border_traffic))

border_state
```
```{r}
#| label: state graph

ggplot(
  border_state,
  aes(x = reorder(state, border_traffic), 
      y = border_traffic,
      fill = state
  )
) +
  geom_col(
    aes(),
    show.legend = FALSE
  ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Amount of Border Traffic per Border State in 2024",
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    x = "State",
    y = "Border Traffic",
    fill = "State"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Analysis:

Texas has by far the most border traffic of any other border state with over 120 million crossings in 2024. Although California is closer behind, there is a steep decrease in border traffic around the Canadian border and other Southern states beyond Texas and California.

### Country Border Traffic

```{r}
#| label: border values

border_country <- clean_border |>
  group_by(border, year) |>
  summarize(
    border_country_traffic = sum(value)
  ) |>
  arrange(year, border_country_traffic)

border_country
```

```{r}
#| label: border graph

ggplot(
  border_country,
  aes(
    x = year,
    y = border_country_traffic,
    color = border
  )
) +
  scale_y_continuous(
    labels = label_comma(),
    limits = c(0, 400000000)
  ) +
  geom_line(
    aes(),
    size = 2
  ) +
  labs(
    title = str_wrap("Border Traffic at the Canadian and Mexican borders from 1996 until 2024"),
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    x = "Year",
    y = "Border Traffic",
    color = "Border"
  ) +
  theme_minimal()
```

Analysis:

Border traffic along the US-Mexico border almost doubled traffic along the US-Canada border. There was a peak for border traffic along the US-Mexico border around and just before 2000, and a dramatic fall in 2020 during the COVID-19 pandemic.

### Transportation Across the Border(s)

```{r}
#| label: transportation

border_transportation <- clean_border |>
  group_by(measure, year) |>
  summarize(
    transportation_numbers = sum(value)
  ) |>
  arrange(year, transportation_numbers)

border_transportation
```
```{r}
#| label: transportation graph

transportation_graph <- ggplot(
  border_transportation,
  aes(
    x = year,
    y = transportation_numbers,
    color = measure
  )
) +
  scale_y_log10(
    breaks = c(100000, 1000000, 10000000, 100000000),
    labels = scales::label_number(accuracy = 1),
  ) +
  geom_line(
    aes(),
    size = 1
  ) +
  labs(
    title = str_wrap("Type of Transportation for Border Traffic from 1996 to 2024"),
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    x = "Year",
    y = "Border Traffic Transportation Amount",
    color = "Mode of Transportation"
  ) +
  theme_minimal()

transportation_graph

ggplotly(transportation_graph)
```

Analysis:

The vast majority of border traffic is personal vehicle traffic. Pedestrian border traffic blew me away with almost 50 million pedestrians crossing the border in 2019. Group transportation took a hit during the pandemic, with buses dropping more than half of their report in 2019, from almost 230,000 to almost 100,000 in 2020.
