---
title: "Border Crossing Entry Data"
---

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
library(ggthemes)
library(scales)
library(plotly)
# Maps
library(mapdata)
library(maps)
library(stringr)
library(viridis)
library(sf)
library(rnaturalearth)
library(ggmagnify)
```

```{r}
#| label: view

show <- read_csv("01-unclean/Border_Crossing_Entry_Data.csv")

show
```


```{r}
#| label: import

border <- read_csv("01-unclean/Border_Crossing_Entry_Data.csv") |>
  clean_names() |>
  mutate(
    border_year = my(date),
    year = year(border_year)
  ) |>
  select(
    port_name,
    state, 
    border, 
    year, 
    measure, 
    value,
    longitude,
    latitude
  )

border
```
### Cleaning

```{r}
#| label: cleaning

border |>
  write_rds("02-clean/border.rds")
```

## Analysis and Visuals

### Numbers by State

```{r}
#| label: clean import

clean_border <- read_rds("02-clean/border.rds")
```

```{r}
#| label: border year

clean_border |>
  count(year)
```


```{r}
#| label: state values

border_state <- clean_border |>
  filter(
    year == 2024
  ) |>
  group_by(state, year) |>
  summarize(
    border_traffic = sum(value)
  ) |>
  mutate(
    state = str_to_upper(state) |>
    str_trim()
  ) |>
  arrange(desc(border_traffic))

border_state
```
```{r}
#| label: state graph

ggplot(
  border_state,
  aes(x = reorder(state, border_traffic), 
      y = border_traffic,
      fill = state
  )
) +
  geom_col(
    aes(),
    show.legend = FALSE
  ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Amount of Border Traffic per Border State in 2024",
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    x = "State",
    y = "Border Traffic",
    fill = "State"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
#| label: county/state coords

usa <- map_data("usa")

states <- map_data("state") |>
  mutate(
    region = str_to_upper(region) |>
    str_trim()
  )
```


```{r}
#| label: ggplot version

usa_plot <- ggplot() +
  geom_polygon(
    data = usa,
    aes(
      x = long,
      y = lat,
      group = group
    ),
    color = "gray2"
  ) +
  coord_quickmap()

states_plot <- usa_plot +
  geom_polygon(
    data = states,
    aes(
      x = long,
      y = lat,
      group = group
    ),
    color = "gray2",
    fill = "white"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),  # Adjust legend text size
    legend.key.width = unit(1, "cm"),  # Widen legend items
    plot.title.position = "plot",  # Align title/subtitle properly
    legend.box = "horizontal",
  ) +
  coord_quickmap()

border_state_graph <- states_plot +
  geom_map(
    data = border_state,
    map = states,
    aes(
      map_id = state,
      fill = border_traffic
    ),
    color = "gray2"
  ) +
  labs(
    title = "Amount of Border Traffic per Border State in 2024",
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    fill = "Border Traffic"
  ) +
  scale_fill_gradient(low = "blue", high = "orange", labels = comma)

border_state_graph
```

Analysis:

Texas has by far the most border traffic of any other border state with over 120 million crossings in 2024. Although California is closer behind, there is a steep decrease in border traffic around the Canadian border and other Southern states beyond Texas and California.

### Country Border Traffic

```{r}
#| label: border values

border_country <- clean_border |>
  group_by(border, year) |>
  summarize(
    border_country_traffic = sum(value)
  ) |>
  arrange(year, border_country_traffic)

border_country
```

```{r}
#| label: border graph

ggplot(
  border_country,
  aes(
    x = year,
    y = border_country_traffic,
    color = border
  )
) +
  scale_y_continuous(
    labels = label_comma(),
    limits = c(0, 400000000)
  ) +
  geom_line(
    aes(),
    size = 2
  ) +
  labs(
    title = str_wrap("Border Traffic at the Canadian and Mexican borders from 1996 until 2024"),
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    x = "Year",
    y = "Border Traffic",
    color = "Border"
  ) +
  theme_minimal()
```

Analysis:

Border traffic along the US-Mexico border almost doubled traffic along the US-Canada border. There was a peak for border traffic along the US-Mexico border around and just before 2000, and a dramatic fall in 2020 during the COVID-19 pandemic.

### Transportation Across the Border(s)

```{r}
#| label: transportation

border_transportation <- clean_border |>
  group_by(measure, year) |>
  summarize(
    transportation_numbers = sum(value)
  ) |>
  arrange(year, transportation_numbers)

border_transportation
```
```{r}
#| label: transportation graph

transportation_graph <- ggplot(
  border_transportation,
  aes(
    x = year,
    y = transportation_numbers,
    color = measure
  )
) +
  scale_y_log10(
    breaks = c(100000, 1000000, 10000000, 100000000),
    labels = scales::label_number(accuracy = 1),
  ) +
  geom_line(
    aes(),
    size = 1
  ) +
  labs(
    title = str_wrap("Type of Transportation for Border Traffic from 1996 to 2024"),
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    x = "Year",
    y = "Border Traffic Transportation Amount",
    color = "Mode of Transportation"
  ) +
  theme_minimal()

transportation_graph

ggplotly(transportation_graph)
```

Analysis:

The vast majority of border traffic is personal vehicle traffic. Pedestrian border traffic blew me away with almost 50 million pedestrians crossing the border in 2019. Group transportation took a hit during the pandemic, with buses dropping more than half of their report in 2019, from almost 230,000 to almost 100,000 in 2020.

```{r}
#| label: coordinate map

border_coord <- border |>
  filter(
    year == "2024",
    state != "Alaska"
  ) |>
  group_by(
    state,
    port_name
  ) |>
  mutate(
    port_count = str_count(unique(port_name, ", "))
  ) |>
  summarize(
    border_count = sum(port_count, na.rm = TRUE),
    value_sum = sum(value, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE), #avg longitude
    latitude = mean(latitude, na.rm = TRUE), #avg latitude
    .groups = "drop"
  ) |>
  arrange(
    desc(value_sum)
  )

border_coord
```

```{r}
#| label: border dot plot

border_coord_plot <- states_plot +
  geom_point(
   data = border_coord,
   alpha = 0.7,
   aes(
     x = longitude,
     y = latitude,
     color = border_count,
     size = value_sum
   )
  ) +
  labs(
    title = str_wrap("Border Traffic Location and Amount in 2024"),
    subtitle = str_wrap("The color of the spots on the map are determined by how many times each port was visited per year. The size of the spots represent the total amount of border crossing in its specific port for all of 2024. Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    color = "Port Visit Instances",
    size = "# of Border Entries"
  ) +
  scale_size_continuous(range = c(1,10), labels = comma) +
  scale_color_gradient(low = "blue", high = "orange", labels = comma)
  

border_coord_plot
```

Revised Analysis:

As reflected by the country border crossing data, the Mexico-America border has much more traffic than the Canada-America border. Most ports are visited between 1,000 and 2,000 times per year, with an exception of many ports in the North. New York sees a lot of border entry instances and quantity of crossing into the United States. Texas and California receive the most border entries of any other state.

```{r}
#| label: coordinate map transportation

border_trans_coord <- border |>
  filter(
    year == "2024",
    state != "Alaska"
  ) |>
  group_by(
    state,
    measure
  ) |>
  mutate(
    transportation_count = str_count(unique(measure, ", "))
  ) |>
  summarize(
    trans_count = sum(transportation_count, na.rm = TRUE),
    value_sum = sum(value, na.rm = TRUE),
    longitude = mean(longitude, na.rm = TRUE), #avg longitude
    latitude = mean(latitude, na.rm = TRUE), #avg latitude
    .groups = "drop"
  ) |>
  arrange(
    desc(value_sum)
  )

border_trans_coord
```

```{r}
#| label: transportation dot plot

border_trans_plot <- states_plot +
  geom_point(
   data = border_trans_coord,
   alpha = 1,
   aes(
     x = longitude,
     y = latitude,
     color = measure,
     size = value_sum
   )
  ) +
  labs(
    title = str_wrap("Border Traffic Location and Amount in 2024"),
    subtitle = str_wrap("The color of the spots on the map are determined by the type of transportation. The size of the spots represent the total amount of border crossing in its specific port for all of 2024. Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: Department of Transportation",
    color = "Type of Transportation",
    size = "Transportation Instances"
  ) +
  scale_size_continuous(range = c(1,10), labels = comma)
  

border_trans_plot
```

Revised Analysis:

Traffic seems to be highly concentrated on a few ports in each border state. There is heavy pedestrian and personal vehicle traffic in Texas, California and New York. States in the Midwest get virtually no traffic in comparison to the South or New York and Washington.