---
title: "Crime in Los Angeles"
format: html
---

## Setup

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(janitor)
library(ggthemes)
library(scales)
library(plotly)
## Map Data
library(mapdata)
library(maps)
library(stringr)
library(viridis)
library(sf)
library(rnaturalearth)
library(ggmagnify)
```


```{r}
#| label: import and clean

raw_data <- read_csv("01-unclean/Los-Angeles_Crime.csv") |>
  clean_names()

la_crime <- raw_data |>
  mutate(
    date_reported = mdy_hms(date_rptd), 
    date_occurred = mdy_hms(date_occ),
    crime_description = crm_cd_desc,
    year_occurred = year(date_occurred),
    latitude_unavg = lat,
    longitude_unavg = lon
  ) |>
  select(
    date_reported,
    date_occurred,
    year_occurred,
    crime_description,
    premis_desc,
    latitude_unavg,
    longitude_unavg,
    area_name
  )

raw_data

la_crime
```

## Analysis and Visualization

### Data Wrapper Attempt -- LA Crime, Location 2024

```{r}
#| label: Location Merge

la_plot <- la_crime |>
  filter(
    year_occurred == "2024"
  ) |>
  group_by(
    area_name,
    crime_description
  ) |>
  mutate(
    area_count = str_count(unique(area_name, ", ")),
    crime_description = str_wrap(crime_description, width = 20)
  ) |>
  summarize(
    incident_count = sum(area_count, na.rm = TRUE),
    longitude = mean(longitude_unavg, na.rm = TRUE),
    latitude = mean(latitude_unavg, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(
    desc(incident_count)
  ) |>
  head(50)
  
la_plot
```


```{r}
#| label: Datawrapper

la_plot |>
  write_csv("02-clean/la-crime.csv")
```

#### State Import

```{r}
#| label: county/state coords

california <- map_data("state") |>
  filter(
    region == "california"
  )

la_county <- map_data("county") |>
  filter(
    region == "california",
    subregion == "los angeles"
  )
```

```{r}
#| label: ggplot version

california_base <- ggplot() +
  geom_polygon(
    data = california,
    aes(
      x = long,
      y = lat,
    ),
    color = "black",
    fill = "cornflowerblue"
  ) +
  coord_quickmap()

cala_plot <- california_base +
  geom_polygon(
    data = la_county,
    aes(
      x = long,
      y = lat,
    ),
    color = "black",
    fill = "white"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),  # Adjust legend text size
    legend.key.width = unit(1, "cm"),  # Widen legend items
    plot.title.position = "plot",  # Align title/subtitle properly
    legend.box = "horizontal"
  ) +
  coord_quickmap()

cala_plot
```
#### Zoom

```{r}
#| label: la plot

la_plot <- ggplot() +
  geom_polygon(
    data = la_county,
    aes(
      x = long,
      y = lat,
    ),
    fill = "white",
    color = "gray"
  ) +
  geom_point(
    data = la_plot,
    alpha = 0.5,
    aes(
      x = longitude,
      y = latitude,
      size = incident_count,
      color = crime_description
    )
  ) +
  scale_size_continuous(range = c(1,10)) +
  coord_cartesian(
      xlim = c(-119, -118.5), 
      ylim = c(33.7, 34.2)
  ) +
  coord_quickmap() +
  labs(
    title = str_wrap("Top 50 Incident Locations in Los Angeles in 2024"),
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: City of Los Angeles",
    x = "Longitude",
    y = "Latitude",
    size = "Incident Count",
    color = "Type of Crime"
  ) +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    legend.title = element_text(size = 8),
    legend.text = element_text(size = 6),  # Adjust legend text size
    legend.key.width = unit(1, "cm"),  # Widen legend items
    plot.title.position = "plot",  # Align title/subtitle properly
    legend.box = "horizontal"
  )

from <- c(xmin = -119, xmax = -118, ymin = 33.7, ymax = 34.5)
# Names xmin, xmax, ymin, ymax are optional:
to <- c(3 - 1, 3 + 1,1, 2) ## UGH NOTHING WORKSSS

la_plot + geom_magnify(from = from, to = to)
```
Map of Los Angeles County: https://maps.app.goo.gl/LNF9MwwGZaaBmdQE6

Revised Analysis:
In 2024, one of the most prevalent crimes in Los Angeles was car theft, with hot spots around downtown LA, Santa Monica and Canoga Park, above Topanga State Park. Santa Monica was also a hot spot for petty and grand theft, around Venice Beach and the Los Angeles airport. Northern LA County, which includes Santa Clarita and Lancaster, is untouched in regards to top 50 areas and type of crime of 2024.

### Type of Crime, Graph

```{r}
#| label: type of crime

crime_year <- la_crime |>
  mutate(
    crime_description = str_wrap(
      crime_description, width = 20
    )
  ) |>
  group_by(
    year_occurred,
    crime_description
  ) |>
  summarize(
    incident_count = n(),
    .groups = "drop"
  ) |>
  group_by(
    year_occurred
  ) |>
  slice_max(
    incident_count,
    n = 10,
    with_ties = FALSE
  )

crime_year
```

```{r}
#| label: type of crime, point graph
#| message: false

top10_plot <- ggplot(
  crime_year,
  aes(
    x = year_occurred,
    y = incident_count,
    color = crime_description
  )
) +
  geom_point(
    show.legend = FALSE,
    size = 2.2
  ) +
  geom_line(
    linewidth = 1.5
  ) +
  labs(
    title = "Top 10 Crimes in Los Angeles from 2020 to 2024",
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: City of Los Angeles",
    x = "Year",
    y = "Number of Incidents",
    color = "Type of Crime"
  )
  theme_minimal()

top10_plot

ggplotly(top10_plot)
```

Analysis:

The most frequent crimes in Los Angeles are non-violent, with an exception of assault and battery. As of 2024, crime in Los Angeles, on all fronts, is declining. Interestingly, in 2022, identity theft was the second most frequent crime, with over 22,000 occurrences. Stolen vehicles have remained at the top of the list over all four years, and it saw a slight peak of almost 25,000 occurrences, also in 2022.

```{r}
#| label: type of crime, bar graph
#| message: false

top10_bar <- ggplot(
  crime_year,
  aes(
    x = crime_description,
    y = incident_count,
    fill = crime_description
  )
) +
  geom_col() +
  labs(
    title = "Top 10 Crimes in Los Angeles from 2020 to 2024",
    subtitle = str_wrap("Chart created by Meredith Roberts for the Data Visualization Independent Study."),
    caption = "Source: City of Los Angeles",
    x = "Year",
    y = "Number of Incidents",
    fill = "Type of Crime"
  ) +
  facet_wrap(
    ~year_occurred
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_blank()
  )

top10_bar

ggplotly(top10_bar)
```

